A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-06 doc/popl2013/intro.tex
intro

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, mutiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can be further decomposed into several levels.   In this paper, we present the operational and static
semantics of Legion, a programming language targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance. A very brief summary
of Legion is that data is organized in a hierarchy of {\em regions}
and subregions and computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying the
operations a task may perform on a region argument (e.g., read or
write, among others) and novel {\em region coherence} annotations that
expresses whether other tasks may perform operations concurrently on
the region (e.g., {\em exclusive} access for task $t$ guarantees no
other task's updates to a region are visible to $t$).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
dependent} (roughly, must be executed in program order to guarantee
the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
the task hierarchy that are {\em siblings} (have the same parent task)
are also dependent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; any centralized scheduler would be a severe bottleneck
because of the huge latencies involved in moving data and computation
in the target class of machines.

% The linked list example
% Add float={t} to the list of options if you want the listing at the top
\begin{lstlisting}[label={lst:linkedlist},caption={Linked List Example}]
type list<r> = < int, list<r>@r >

@@

The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
of integers, illustrates region {\em partitions}, another important
building block of Legion programs.  The region arguments to {\tt
make\_tree} are {\tt rl}, the region holding all list elements, {\tt
rls}, the subregion of {\tt rl} holding those list elements to be
indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
hold the tree itself.  Line 21 divides the list elements in {\tt rls}
into those that are less than and those that are greater than a split
value (which is arbitrarily chosen to be the value of the first
element of the list).  Partitioning is done by constructing a {\em
coloring} using the task {\tt color\_list} (line 6), which traverses the list
and for each list element $e$ in the subregion (called {\tt rs} in
{\tt color\_region}) either colors the element 1 or 2 (colors are
integers) depending on its relationship to the split value.  A
coloring is just a distinghuished function, built using the Legion
primitive {\tt color}, that maps elements of a region to integers.
When used in a {\tt partition} expression the coloring defines how to
populate new subregtions of the partitioned region.  On line 22, the
list elements in {\tt rls} are partitioned into two subregions {\tt
rll} and {\tt rlr} of list elements less than, and greater than or
equal to, the split value, respectively.  Another use of {\tt
partition} is seen on line 21 where the {\tt rt} region is partitioned
into subregions to hold the left {\tt rlt} and right {\tt rtr}
subtrees, respectively.  Because region {\tt rt} is empty at this
point (has no elements), the empty coloring function {\tt newcolor rt}
for the region {\tt rt} is given to the partitioning operation; the
two new subregions are also empty.  These examples illustrate that
partitioning can be used for two distinct, and equally important,
styles of computation: {\em allocate-then-partition}, in which a
global collection of data is created an then divided for
(usually parallel) processing, and {\em partition-then-allocate}, in
which a collection of empty but disjoint subregions is created and
then populated with data on which (usually parallel) processing will
take place.

Lines 26-27 illustrates {\em packing} a region relationship into the heap.
Regions are first-class values in Legion and can be stored in data structures.
Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
stored with the tree value.  The type system guarantees statically that the packed value
satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
the packed value is stored.  When the region relationship value is read 
from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are known to 
satisfy these constraints.  A key design decision in Legion, one that turns out to be
central to all the soundness results as well as the ability to perform distributed, parallel
scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
are not captured by region relationships.  When a region is stored in the heap it has no
permissions or coherence associated with it; these belong to the task.  When the region relationship
is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
write permission for region {\tt rt} and $\tt rtl \leq rt$ is a constraint of the region relationship; holding
permission for a region $r$ implies the same permission for any subregion of $r$ (note
that the region {\tt rt} is not one of the existentially quantified regions).

@@

The last example, shown in Listing~\ref{lst:histogram}, is a histogram
computation that recursively walks a {\tt tree}.  The first version 
uses a {\em coherence annotation} to relax the
required ordering of tasks.  When there is no coherence annotation on
a region $r$ used by a task $t$, the task implicitly requires $r$
with {\em exclusive} coherence, meaning that any other task $t'$ that might
update $r$ in a way visible to $t$ must appear to execute in normal sequential
order with respect to $t$.  The task {\tt hist\_tree1} only requires the region {\tt rh}
with {\tt atomic} coherence, meaning that other tasks that also use {\tt rh} with {\tt atomic} coherence
may be reordered to appear to execute atomically before or after the task.  The task {\tt hist\_tree2} achieves
a similar affect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
with the same associative-commutative operator to be reordered.

>>>>

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, mutiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can be further decomposed into several levels.   In this paper, we present the operational and static
semantics of Legion, a programming model targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance.

Briefly, in Legion data is organized in a hierarchy of {\em regions}
and subregions and computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying the
operations a task may perform on a region argument (read,
write, or reduce) and novel {\em region coherence} annotations that
express whether other tasks may perform operations concurrently on
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
dependent} (roughly, must be executed in program order to guarantee
the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
the task hierarchy that are {\em siblings} (have the same parent task)
are also dependent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in moving data and computation
in the target class of machines.

% The linked list example
% Add float={t} to the list of options if you want the listing at the top
\begin{lstlisting}[float={t},label={lst:linkedlist},caption={Linked List Example}]
type list<r> = < int, list<r>@r >

@@

The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
of integers, illustrates region {\em partitions}, another important
building block of Legion programs.  The region arguments to {\tt
make\_tree} are {\tt rl}, the region holding all list elements, {\tt
rls}, the subregion of {\tt rl} holding those list elements to be
indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
hold the tree itself.  Line 21 divides the list elements in {\tt rls}
into those that are less than and those that are greater than a split
value (which is arbitrarily chosen to be the value of the first
element of the list).  Partitioning is done by constructing a {\em
coloring} using the task {\tt color\_list} (line 6), which traverses the list
and for each list element in the subregion (region {\tt rs} in
{\tt color\_region}) either colors the element 1 or 2 (colors are
integers) depending on its relationship to the split value.  A
coloring is just a distinghuished function, built using the Legion
primitive {\tt color}, that maps elements of a region to integers.
When used in a {\tt partition} expression the coloring defines how to
populate new subregions of the partitioned region.  On line 22, the
list elements in {\tt rls} are partitioned into two subregions {\tt
rll} and {\tt rlr} of list elements less than, and greater than or
equal to, the split value, respectively.  Another use of {\tt
partition} is seen on line 21 where the {\tt rt} region is partitioned
into subregions to hold the left {\tt rlt} and right {\tt rtr}
subtrees, respectively.  Because region {\tt rt} is empty at this
point (has no elements), the empty coloring function {\tt newcolor rt}
for the region {\tt rt} is given to the partitioning operation; the
two new subregions are also empty.  These examples illustrate that
partitioning can be used for two distinct, and equally important,
styles of computation: {\em allocate-then-partition}, in which a
global collection of data is created and then subdivided for
(usually parallel) processing, and {\em partition-then-allocate}, in
which a collection of empty but disjoint subregions is created and
then populated with data on which (usually parallel) processing will
take place.

Lines 26-27 illustrates {\em packing} a region relationship into the heap.
Regions are first-class values in Legion and can be stored in data structures.
Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
stored with the tree value.  The type system guarantees statically that the packed value
satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
the packed value is stored.  When the region relationship value is read 
from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are known to 
satisfy these constraints.  A key design decision in Legion, one that turns out to be
central to the soundness results as well as the ability to perform distributed, parallel
scheduling of tasks, is that permissions and coherence annotations 
are not captured by region relationships.  When a region is stored in the heap it has no
permissions or coherence associated with it; these belong to the task.  When the region relationship
is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
write permission for region {\tt rt} and $\tt rtl \leq rt$ is a constraint of the region relationship; holding
permission for a region $r$ implies the same permission for any subregion of $r$ (note
that the region {\tt rt} is not one of the existentially quantified regions).

@@

The last example, shown in Listing~\ref{lst:histogram}, is a histogram
computation that recursively walks a {\tt tree}.  The first version 
uses a {\em coherence annotation} to relax the
required ordering of tasks.  When there is no coherence annotation on
a region $r$ used by a task $t$, the task implicitly requires $r$
with {\em exclusive} coherence, meaning that any other task $t'$ that might
update $r$ in a way visible to $t$ must appear to execute in normal sequential
order with respect to $t$.  The task {\tt hist\_tree1} only requires the region {\tt rh}
with {\tt atomic} coherence, meaning that other tasks that also use {\tt rh} with {\tt atomic} coherence
may be reordered to appear to execute atomically before or after the task.  The task {\tt hist\_tree2} achieves
a similar effect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
with the same associative-commutative operator to be reordered.
